---
title: "COMPUTER LAB 1 Group report"
author: "Andrea Bruzzone,Thomas Zhang"
date: "2016 M02 3"
output: pdf_document
---

##Assignment 1
The result of the program wrote by a student is:

```{r}
x1 <- 1/3
x2 <- 1/4
if(x1-x2 == 1/12){
  print("Teacher said true")
}else{
  print("Teacher lied")
}
```

However, the result should be "Teacher said true", since 1/3 - 1/4 = 1/12. I imagine this is due to the finite nature of the significand of floating point numbers.

In order to get a correct result, we should round both the difference $x1 - x2$ and 1/12 with the same decimal places, it seems that using less or equal than 16 decimal places the result is correct. We chose to round with 4 decimal places.

The correct code should be like:

```{r}
x1 <- 1/3
x2 <- 1/4
if(round(x1-x2, 4)==round(1/12, 4)){
  print("Teacher said true")
}else{
  print("Teacher lied")
}
```


## Assignment 2

Using:
$$f'(x) = \frac{f(x + \epsilon) - f(x)}{\epsilon}$$
we compute the derivative of f(x) = x. 

For `x = 100000` and `epsilon = 10^-15`,we implement this in R below.
```{r,echo=TRUE}
derivative <- function(x,epsilon){
  res <- ((x + epsilon) - x) / epsilon 
  return(res)
}

derivative(x = 100000, epsilon = 10^-15)
```

The obtained value is zero, while the real value is one. We have a wrong value because $\epsilon$ is much smaller than x, so $x + \epsilon$ give a result which is rounded to x in floating-point arithmetics and so the numerator is 0, resulting in the derivative to be 0.

When we try a larger epsilon (a smaller difference in magnitude between $x + \epsilon$ and $x$) the obtained value will be one. The numerical breakdown happens when order difference is around 16.

```{r}
derivative(x = 100000, epsilon = 10^-5)
```

##Assignment 3
We write a function myvar estimating the variance, generate a vector $x = (x_1,.., x_{10000})$ with 10000 random numbers, normally distributed with mean $10^8$ and variance 1.
For each subset $X_i= (x_1,.., x_i)$ with $i = 1...10000$ we compute the difference $Y_i$ between our function and the R function var.

We plot this difference against the dimension of the subset:

```{r, echo=FALSE}
set.seed(12345)
x <- rnorm(10000, mean = 10^8, sd = 1)
myvar <- function(x){
  v <- 1/(length(x) - 1) * (sum(x^2) - 1/length(x) * (sum(x)^2))
  return(v)
}

y <- c()
for(i in 1:10000){
  sub <- x[1:i]
  y[i] <- myvar(sub) - var(sub)
}

plot(1:10000, y, xlab = "Dimension of x", ylab = "Y")
```

From the plot we can see that the difference is never zero so `myvar` and the function `var` never give the same value. The reason could be catastrophic cancellation, since we are adding two numbers with almost equal magnitude but opposite sign, this brings a high probability of roundoff error, considering the narrowness of the standard deviation of the $x_{i}s$, (They are $\mathcal{N}(10^8,1)$ distributed).

Moreover, if we print the results for the subset $X_i= (x_1,.., x_i)$ with $i = 1...10$ for the `myvar` function and the `var` function respectively we see that our function works quite bad and this could be due to the catastrophic cancelltion problem.

`myvar`: 
```{r, echo=FALSE}
for(i in 2:10){
print(myvar(x[1:i]))
}
```

`var`:
```{r, echo=FALSE}
for(i in 2:10){
print(var(x[1:i]))
}  
```

## Assignment 4

We want to solve, in the sense of least squares the  linear regression system $A\beta=b$ where $A=X'X$ are derived from predictors from the `tecator.xls` data sheet and $b=X'Y$ is derived from one column chosen as response. We will try to use the built-in `solve` function.

```{r,echo=TRUE,eval=FALSE}
X <- as.matrix(data[,-c(1,102)])
#head(X)
Y <- as.matrix(data[,102,drop=FALSE])
A <- t(X) %*% X
b <- t(X) %*% Y
solveres <- solve(A,b) # can not solve the system is ill-conditioned
rcond(A) # rcond returns 7.25 * 10^(-17)
```

Due to at least two columns in the $X$ matrix being almost linearly dependent, the condition number, the inverse of the number produced by `rcond`, of `A` is large.
Since the `solve` function seems not to work because the condition number of `A` is too large, we will try to scale the data in hope it will reduce condition number and make `solve` work.

```{r,echo=TRUE,eval=FALSE}
datascaled <- scale(data)

#head(data)
Xscaled <- as.matrix(datascaled[,-c(1,102)])
#head(X)
Yscaled<- as.matrix(datascaled[,102,drop=FALSE])
A <- t(Xscaled) %*% Xscaled
b <- t(Xscaled) %*% Yscaled
solveres <- solve(A,b) #now it works. scaling appears to improve rcond a little
rcond(A) # 7.0 * 10^(-14)
solveres
```

The condition number is smaller now and we are lucky that it works now. Generally we should probably decompose the matrix `A` using QR or SVD or Cholesky decomposition/factorization.

##Appendix

###R code
```{r code=readLines(knitr::purl("C:/Users/Dator/Documents/R_HW/CompStats/GroupLabReport1.Rmd", documentation = 0)), eval = FALSE}
